---
title: "Introduction to betaregscale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to betaregscale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Overview

The **betaregscale** package provides maximum-likelihood estimation of beta
regression models for responses derived from bounded rating scales.
Common examples include pain intensity scales (NRS-11, NRS-21, NRS-101),
Likert-type scales, product quality ratings, and any instrument whose
response can be mapped to the open interval $(0, 1)$.

The key idea is that a discrete score recorded on a bounded scale carries
measurement uncertainty inherent to the instrument. For instance, a pain
score of $y = 6$ on a 0--10 NRS is not an exact value but rather
represents a range: after rescaling to $(0, 1)$, the observation is
treated as interval-censored in $[0.55, 0.65]$. The package uses the
beta distribution to model such data, building a complete likelihood that
supports mixed censoring types within the same dataset.

## Installation

```{r install, eval = FALSE}
# Development version from GitHub:
# install.packages("remotes")
remotes::install_github("evandeilton/betaregscale")
```

```{r library}
library(betaregscale)
```

## Censoring types

The complete likelihood (Lopes, 2024, Eq. 2.24) supports four censoring
types, automatically classified by `check_response()`:

| $\delta$ | Type | Likelihood contribution |
|:--------:|:-----|:-----------------------|
| 0 | Exact (uncensored) | $f(y_i;\, a_i, b_i)$ |
| 1 | Left-censored ($y = 0$) | $F(u_i;\, a_i, b_i)$ |
| 2 | Right-censored ($y = K$) | $1 - F(l_i;\, a_i, b_i)$ |
| 3 | Interval-censored | $F(u_i;\, a_i, b_i) - F(l_i;\, a_i, b_i)$ |

where $f(\cdot)$ and $F(\cdot)$ are the beta density and CDF,
$[l_i, u_i]$ are the interval endpoints, and $(a_i, b_i)$ are the beta
shape parameters derived from $\mu_i$ and $\phi_i$ via the chosen
reparameterization.

## Interval construction

Scale observations are mapped to $(0, 1)$ with uncertainty intervals
controlled by the `type` argument:

- `"m"` (midpoint): $y_t = y/K$, interval $[y_t - 0.5/K,\; y_t + 0.5/K]$
- `"l"` (left-aligned): $y_t = y/K$, interval $[y_t,\; y_t + 1/K]$
- `"r"` (right-aligned): $y_t = y/K$, interval $[y_t - 1/K,\; y_t]$

where $K$ is the number of scale categories (`ncuts`).

```{r check-response}
# Illustrate check_response with a 0-10 NRS scale
y_example <- c(0, 3, 5, 7, 10)
cr <- check_response(y_example, type = "m", ncuts = 10)
cr
```

The `delta` column shows that $y = 0$ is left-censored ($\delta = 1$),
$y = 10$ is right-censored ($\delta = 2$), and all interior values are
interval-censored ($\delta = 3$).

## Data preparation with `bs_prepare()`

In practice, analysts may want to supply their own censoring indicators
or interval endpoints rather than relying on the automatic classification
of `check_response()`. The `bs_prepare()` function provides a flexible,
validated bridge between raw analyst data and `betaregscale()`.

It supports four input modes:

### Mode 1: Score only (automatic)

```{r bs-prepare-mode1}
# Equivalent to check_response <U+2014> delta inferred from y
d1 <- data.frame(y = c(0, 3, 5, 7, 10), x1 = rnorm(5))
bs_prepare(d1, ncuts = 10)
```

### Mode 2: Score + explicit censoring indicator

```{r bs-prepare-mode2}
# Analyst specifies delta directly
d2 <- data.frame(
  y     = c(50, 0, 99, 50),
  delta = c(0, 1, 2, 3),
  x1    = rnorm(4)
)
bs_prepare(d2, ncuts = 100)
```

### Mode 3: Interval endpoints with NA patterns

When the analyst provides `left` and/or `right` columns, censoring is
inferred from the NA pattern:

```{r bs-prepare-mode3}
d3 <- data.frame(
  left  = c(NA, 20, 30, NA),
  right = c(5, NA, 45, NA),
  y     = c(NA, NA, NA, 50),
  x1    = rnorm(4)
)
bs_prepare(d3, ncuts = 100)
```

### Mode 4: Analyst-supplied intervals

When the analyst provides `y`, `left`, and `right` simultaneously, their
endpoints are used directly (rescaled by $K$):

```{r bs-prepare-mode4}
d4 <- data.frame(
  y     = c(50, 75),
  left  = c(48, 73),
  right = c(52, 77),
  x1    = rnorm(2)
)
bs_prepare(d4, ncuts = 100)
```

### Using prepared data with `betaregscale()`

Data processed by `bs_prepare()` is automatically detected by
`betaregscale()` — the internal `check_response()` step is skipped:

```{r bs-prepare-workflow}
set.seed(42)
n <- 200
dat <- data.frame(x1 = rnorm(n), x2 = rnorm(n))
sim <- betaregscale_simulate(
  formula = ~ x1 + x2, data = dat,
  beta = c(0.2, -0.5, 0.3), phi = 1 / 5
)
prep <- bs_prepare(sim, ncuts = 100)
fit_prep <- betaregscale(y ~ x1 + x2, data = prep)
summary(fit_prep)
```

## Example 1: Fixed dispersion model

### Simulating data

We simulate 200 observations from a beta regression model with fixed
dispersion, two covariates, and logit link for the mean.

```{r sim-fixed}
set.seed(4255)
n <- 200
dat <- data.frame(x1 = rnorm(n), x2 = rnorm(n))

sim_fixed <- betaregscale_simulate(
  formula  = ~ x1 + x2,
  data     = dat,
  beta     = c(0.3, -0.6, 0.4),
  phi      = 1 / 10,
  link     = "logit",
  link_phi = "logit",
  ncuts    = 100,
  type     = "m",
  repar    = 2
)

head(sim_fixed)
```

The `type = "m"` argument means that each observation is centered in its
interval. For example, a score of 67 on a 0--100 scale yields
$y_t = 0.67$ with interval $[0.665, 0.675]$.

### Fitting the model

```{r fit-fixed}
fit_fixed <- betaregscale(
  y ~ x1 + x2,
  data     = sim_fixed,
  link     = "logit",
  link_phi = "logit",
  repar    = 2
)
summary(fit_fixed)
```

The summary output follows the `betareg` package style, showing separate
coefficient tables for the mean and precision submodels, with Wald
z-tests and $p$-values based on the standard normal distribution.

### Goodness of fit

```{r gof-fixed}
gof(fit_fixed)
```

### Comparing link functions

The package supports several link functions for the mean submodel. We
can compare them using information criteria:

```{r compare-links}
links <- c("logit", "probit", "cauchit", "cloglog")
fits <- lapply(setNames(links, links), function(lnk) {
  betaregscale(y ~ x1 + x2, data = sim_fixed, link = lnk, repar = 2)
})

# Estimates
est_table <- do.call(rbind, lapply(names(fits), function(lnk) {
  e <- est(fits[[lnk]])
  e$link <- lnk
  e
}))
est_table

# Goodness of fit
gof_table <- do.call(rbind, lapply(fits, gof))
gof_table
```

### Residual diagnostics

The `plot()` method provides six diagnostic panels. By default, the
first four are shown:

```{r plot-fixed, fig.height = 6}
plot(fit_fixed)
```

For ggplot2 output (requires the **ggplot2** package):

```{r plot-fixed-gg, eval = requireNamespace("ggplot2", quietly = TRUE), fig.height = 6}
plot(fit_fixed, gg = TRUE)
```

### Predictions

```{r predict-fixed}
# Fitted means
head(predict(fit_fixed, type = "response"))

# Conditional variance
head(predict(fit_fixed, type = "variance"))

# Quantile predictions
head(predict(fit_fixed, type = "quantile", at = c(0.25, 0.5, 0.75)))
```

### Confidence intervals

Wald confidence intervals based on the asymptotic normal approximation:

```{r confint-fixed}
confint(fit_fixed)
confint(fit_fixed, model = "mean")
```

### Censoring structure

The `censoring_summary()` function provides a visual and tabular
overview of the censoring types in the fitted model:

```{r censoring-summary, fig.height = 5}
censoring_summary(fit_fixed)
```

## Example 2: Variable dispersion model

In many applications, the dispersion parameter $\phi$ may depend on
covariates. The package supports variable-dispersion models using the
`Formula` package notation: `y ~ x1 + x2 | z1 + z2`, where the terms
after `|` define the linear predictor for $\phi$.

### Simulating data

```{r sim-variable}
set.seed(2222)
n <- 200
dat_z <- data.frame(
  x1 = rnorm(n),
  x2 = rnorm(n),
  x3 = rbinom(n, size = 1, prob = 0.5),
  z1 = rnorm(n),
  z2 = rnorm(n)
)

sim_var <- betaregscale_simulate_z(
  formula_x = ~ x1 + x2 + x3,
  formula_z = ~ z1 + z2,
  data = dat_z,
  beta = c(0.2, -0.6, 0.2, 0.2),
  zeta = c(0.2, -0.8, 0.6),
  link = "logit",
  link_phi = "logit",
  ncuts = 100,
  type = "m",
  repar = 2
)

head(sim_var)
```

### Fitting the model

```{r fit-variable}
fit_var <- betaregscale(
  y ~ x1 + x2 | z1,
  data     = sim_var,
  link     = "logit",
  link_phi = "logit",
  repar    = 2
)
summary(fit_var)
```

Notice the `(phi)_` prefix in the precision coefficient names, following
the `betareg` convention.

### Accessing coefficients by submodel

```{r coef-variable}
# Full parameter vector
coef(fit_var)

# Mean submodel only
coef(fit_var, model = "mean")

# Precision submodel only
coef(fit_var, model = "precision")

# Variance-covariance matrix for the mean submodel
vcov(fit_var, model = "mean")
```

### Comparing link functions (variable dispersion)

```{r compare-links-var}
links <- c("logit", "probit", "cauchit", "cloglog")
fits_var <- lapply(setNames(links, links), function(lnk) {
  betaregscale(y ~ x1 + x2 | z1, data = sim_var, link = lnk, repar = 2)
})

# Estimates
est_var <- do.call(rbind, lapply(names(fits_var), function(lnk) {
  e <- est(fits_var[[lnk]])
  e$link <- lnk
  e
}))
est_var

# Goodness of fit
gof_var <- do.call(rbind, lapply(fits_var, gof))
gof_var
```

### Diagnostics for variable dispersion

```{r plot-variable, fig.height = 6}
plot(fit_var)
```

## S3 methods reference

The following standard S3 methods are available for objects of class
`"betaregscale"`:

| Method | Description |
|:-------|:-----------|
| `print()` | Compact display of call and coefficients |
| `summary()` | Detailed output with Wald tests and goodness-of-fit |
| `coef(model=)` | Extract coefficients (full, mean, or precision) |
| `vcov(model=)` | Variance-covariance matrix (full, mean, or precision) |
| `confint(model=)` | Wald confidence intervals |
| `logLik()` | Log-likelihood value |
| `AIC()`, `BIC()` | Information criteria |
| `nobs()` | Number of observations |
| `formula()` | Model formula |
| `model.matrix(model=)` | Design matrix (mean or precision) |
| `fitted()` | Fitted mean values |
| `residuals(type=)` | Residuals: response, pearson, rqr, weighted, sweighted |
| `predict(type=)` | Predictions: response, link, precision, variance, quantile |
| `plot(gg=)` | Diagnostic plots (base R or ggplot2) |

## Reparameterizations

The package supports three reparameterizations of the beta distribution,
controlled by the `repar` argument:

**Direct (`repar = 0`):** Shape parameters $a = \mu$ and $b = \phi$
are used directly. This is rarely used in practice.

**Precision (`repar = 1`, Ferrari & Cribari-Neto, 2004):** The mean
$\mu \in (0,1)$ and precision $\phi > 0$ yield $a = \mu\phi$ and
$b = (1-\mu)\phi$. Higher $\phi$ means less variability.

**Mean--variance (`repar = 2`):** The mean $\mu \in (0,1)$ and
dispersion $\phi \in (0,1)$ yield $a = \mu(1-\phi)/\phi$ and
$b = (1-\mu)(1-\phi)/\phi$. Here $\phi$ acts as a coefficient of
variation: smaller $\phi$ means less variability.

```{r reparam}
# Precision parameterization: mu = 0.5, phi = 10 (high precision)
beta_reparam(mu = 0.5, phi = 10, repar = 1)

# Mean-variance parameterization: mu = 0.5, phi = 0.1 (low dispersion)
beta_reparam(mu = 0.5, phi = 0.1, repar = 2)
```

## References

- Lopes, J. E. (2024). *Beta Regression for Interval-Censored
  Scale-Derived Outcomes*. MSc Dissertation, PPGMNE/UFPR.

- Ferrari, S. and Cribari-Neto, F. (2004). Beta regression for
  modelling rates and proportions. *Journal of Applied Statistics*,
  **31**(7), 799--815.
