---
title: "Mixed-Effects Beta Interval Regression with brsmm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Effects Beta Interval Regression with brsmm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  digits = 4
)
```

## Overview

`brsmm()` extends `brs()` to clustered data by adding a Gaussian random
intercept in the mean submodel while preserving the interval-censored
beta likelihood for scale-derived outcomes.

This vignette covers:

1. full model mathematics;
2. estimation by marginal maximum likelihood (Laplace approximation);
3. practical use of all current `brsmm` methods;
4. inferential and validation workflows, including parameter recovery.

```{r library}
library(betaregscale)
```

## Mathematical model

Assume observations \(i = 1, \dots, n_j\) within groups
\(j = 1, \dots, G\), with random intercept \(b_j \sim \mathcal{N}(0,\sigma_b^2)\).

### Linear predictors

\[
\eta_{\mu,ij} = x_{ij}^\top \beta + b_j,
\qquad
\eta_{\phi,ij} = z_{ij}^\top \gamma
\]

\[
\mu_{ij} = g^{-1}(\eta_{\mu,ij}),
\qquad
\phi_{ij} = h^{-1}(\eta_{\phi,ij})
\]

with \(g(\cdot)\) and \(h(\cdot)\) chosen by `link` and `link_phi`.

### Beta parameterization

For each \((\mu_{ij}, \phi_{ij})\), `repar` maps to beta shape parameters
\((a_{ij}, b_{ij})\) via `brs_repar()`.

### Conditional contribution by censoring type

Each observation contributes:

\[
L_{ij}(b_j;\theta)=
\begin{cases}
f(y_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=0\\
F(u_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=1\\
1 - F(l_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=2\\
F(u_{ij}; a_{ij}, b_{ij}) - F(l_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=3
\end{cases}
\]

where \(l_{ij},u_{ij}\) are interval endpoints on \((0,1)\),
\(f(\cdot)\) is beta density, and \(F(\cdot)\) is beta CDF.

### Group marginal likelihood

\[
L_j(\theta)=\int_{\mathbb{R}}
\prod_{i=1}^{n_j} L_{ij}(b_j;\theta)\,
\varphi(b_j;0,\sigma_b^2)\,db_j
\]

\[
\ell(\theta)=\sum_{j=1}^G \log L_j(\theta)
\]

### Laplace approximation used by `brsmm()`

Define
\[
Q_j(b)=\sum_{i=1}^{n_j}\log L_{ij}(b;\theta)+\log\varphi(b;0,\sigma_b^2)
\]
and \(\hat b_j=\arg\max_b Q_j(b)\), with curvature
\(H_j=-Q_j''(\hat b_j)\). Then

\[
\log L_j(\theta) \approx
Q_j(\hat b_j) + \frac{1}{2}\log(2\pi) - \frac{1}{2}\log(H_j).
\]

`brsmm()` maximizes the approximated \(\ell(\theta)\) with
`stats::optim()`, and computes group-level posterior modes
\(\hat b_j\) and local standard deviations from the local curvature.

## Simulating clustered scale data

The next helper simulates data from a known mixed model to illustrate fitting,
inference, and recovery checks.

```{r simulate-data}
sim_brsmm_data <- function(seed = 3501L, g = 24L, ni = 12L,
                           beta = c(0.20, 0.65),
                           gamma = c(-0.15),
                           sigma_b = 0.55) {
  set.seed(seed)
  id <- factor(rep(seq_len(g), each = ni))
  n <- length(id)
  x1 <- rnorm(n)

  b_true <- rnorm(g, mean = 0, sd = sigma_b)
  eta_mu <- beta[1] + beta[2] * x1 + b_true[as.integer(id)]
  eta_phi <- rep(gamma[1], n)

  mu <- plogis(eta_mu)
  phi <- plogis(eta_phi)
  shp <- brs_repar(mu = mu, phi = phi, repar = 2)
  y <- round(stats::rbeta(n, shp$shape1, shp$shape2) * 100)

  list(
    data = data.frame(y = y, x1 = x1, id = id),
    truth = list(beta = beta, gamma = gamma, sigma_b = sigma_b, b = b_true)
  )
}

sim <- sim_brsmm_data()
str(sim$data)
```

## Fitting `brsmm()`

```{r fit-brsmm}
fit_mm <- brsmm(
  y ~ x1,
  random = ~ 1 | id,
  data = sim$data,
  repar = 2,
  int_method = "laplace",
  method = "BFGS",
  control = list(maxit = 1000)
)

summary(fit_mm)
```

## Core methods

### Coefficients and random effects

`coef(fit_mm, model = "random")` is on the optimizer scale
(\(\log \sigma_b\)); use `exp()` to report \(\sigma_b\).

```{r methods-coef}
knitr::kable(
  data.frame(parameter = names(coef(fit_mm, model = "full")),
             estimate = as.numeric(coef(fit_mm, model = "full"))),
  digits = 4
)
knitr::kable(
  data.frame(log_sigma_b = as.numeric(coef(fit_mm, model = "random")),
             sigma_b = as.numeric(exp(coef(fit_mm, model = "random")))),
  digits = 4
)
knitr::kable(head(ranef.brsmm(fit_mm)), digits = 4)
```

### Variance-covariance, summary and likelihood criteria

```{r methods-summary}
vc <- vcov(fit_mm)
dim(vc)

sm <- summary(fit_mm)
knitr::kable(sm$coefficients, digits = 4)

knitr::kable(
  data.frame(
    logLik = as.numeric(logLik(fit_mm)),
    AIC = AIC(fit_mm),
    BIC = BIC(fit_mm),
    nobs = nobs(fit_mm)
  ),
  digits = 4
)
```

### Fitted values, prediction and residuals

```{r methods-predict}
knitr::kable(
  data.frame(
    mu_hat = head(fitted(fit_mm, type = "mu")),
    phi_hat = head(fitted(fit_mm, type = "phi")),
    pred_mu = head(predict(fit_mm, type = "response")),
    pred_eta = head(predict(fit_mm, type = "link")),
    pred_phi = head(predict(fit_mm, type = "precision")),
    pred_var = head(predict(fit_mm, type = "variance"))
  ),
  digits = 4
)

knitr::kable(
  data.frame(
    res_response = head(residuals(fit_mm, type = "response")),
    res_pearson = head(residuals(fit_mm, type = "pearson"))
  ),
  digits = 4
)
```

### Diagnostic plotting methods

`plot.brsmm()` supports base and ggplot2 backends:

```{r methods-plot}
plot(fit_mm, which = 1:4, type = "pearson")

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot(fit_mm, which = 1:2, gg = TRUE)
}
```

`autoplot.brsmm()` provides focused ggplot diagnostics:

```{r methods-autoplot}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  autoplot.brsmm(fit_mm, type = "calibration")
  autoplot.brsmm(fit_mm, type = "score_dist")
  autoplot.brsmm(fit_mm, type = "ranef_qq")
  autoplot.brsmm(fit_mm, type = "residuals_by_group")
}
```

### Prediction with `newdata`

If `newdata` contains unseen groups, `predict.brsmm()` uses random effect
equal to zero for those levels.

```{r methods-newdata}
nd <- sim$data[1:8, c("x1", "id")]
knitr::kable(
  data.frame(pred_seen = as.numeric(predict(fit_mm, newdata = nd, type = "response"))),
  digits = 4
)

nd_unseen <- nd
nd_unseen$id <- factor(rep("new_cluster", nrow(nd_unseen)))
knitr::kable(
  data.frame(pred_unseen = as.numeric(predict(fit_mm, newdata = nd_unseen, type = "response"))),
  digits = 4
)
```

## Statistical tests and validation workflow

### Wald tests (from `summary`)

`summary.brsmm()` reports Wald \(z\)-tests for each parameter:
\[
z_k = \hat\theta_k / \mathrm{SE}(\hat\theta_k).
\]

```{r wald-tests}
sm <- summary(fit_mm)
knitr::kable(sm$coefficients, digits = 4)
```

### Likelihood-ratio test for nested fixed effects

For nested models with the same random-effect structure, an LR test can be
computed as:
\[
LR = 2\{\ell(\hat\theta_1)-\ell(\hat\theta_0)\},
\]
using a \(\chi^2\) reference with parameter-count difference.

```{r lr-test, eval = FALSE}
fit_null <- brsmm(
  y ~ 1,
  random = ~ 1 | id,
  data = sim$data,
  repar = 2,
  int_method = "laplace",
  method = "BFGS",
  control = list(maxit = 1000)
)

ll0 <- logLik(fit_null)
ll1 <- logLik(fit_mm)
lr_stat <- 2 * (as.numeric(ll1) - as.numeric(ll0))
df_diff <- attr(ll1, "df") - attr(ll0, "df")
p_value <- stats::pchisq(lr_stat, df = df_diff, lower.tail = FALSE)

c(LR = lr_stat, df = df_diff, p_value = p_value)
```

### Residual diagnostics (quick checks)

```{r quick-diagnostics}
r <- residuals(fit_mm, type = "pearson")
knitr::kable(
  data.frame(
    mean = mean(r),
    sd = stats::sd(r),
    q025 = as.numeric(stats::quantile(r, 0.025)),
    q975 = as.numeric(stats::quantile(r, 0.975))
  ),
  digits = 4
)
```

## Parameter recovery experiment

A single-fit recovery table can be produced directly from the previous fit:

```{r recovery-single}
est <- c(
  beta0 = coef(fit_mm, model = "mean")[1],
  beta1 = coef(fit_mm, model = "mean")[2],
  sigma_b = exp(coef(fit_mm, model = "random"))
)

true <- c(
  beta0 = sim$truth$beta[1],
  beta1 = sim$truth$beta[2],
  sigma_b = sim$truth$sigma_b
)

recovery_table <- data.frame(
  parameter = names(true),
  true = as.numeric(true),
  estimate = as.numeric(est[names(true)]),
  bias = as.numeric(est[names(true)] - true)
)
knitr::kable(recovery_table, digits = 4)
```

For a Monte Carlo recovery study, repeat simulation and fitting across
replicates:

```{r recovery-montecarlo, eval = FALSE}
mc_recovery <- function(R = 50L, seed = 7001L) {
  set.seed(seed)
  out <- vector("list", R)

  for (r in seq_len(R)) {
    sim_r <- sim_brsmm_data(seed = seed + r)
    fit_r <- brsmm(
      y ~ x1,
      random = ~ 1 | id,
      data = sim_r$data,
      repar = 2,
      int_method = "laplace",
      method = "BFGS",
      control = list(maxit = 1000)
    )

    out[[r]] <- c(
      beta0 = coef(fit_r, model = "mean")[1],
      beta1 = coef(fit_r, model = "mean")[2],
      sigma_b = exp(coef(fit_r, model = "random"))
    )
  }

  est <- do.call(rbind, out)
  truth <- c(beta0 = 0.20, beta1 = 0.65, sigma_b = 0.55)

  data.frame(
    parameter = colnames(est),
    truth = as.numeric(truth[colnames(est)]),
    mean_est = colMeans(est),
    bias = colMeans(est) - truth[colnames(est)],
    rmse = sqrt(colMeans((sweep(est, 2, truth[colnames(est)], "-"))^2))
  )
}

mc_recovery(R = 50)
```

## How this maps to automated package tests

The package test suite includes dedicated `brsmm` tests for:

1. fitting with Laplace integration;
2. one- and two-part formulas;
3. S3 methods (`coef`, `vcov`, `summary`, `predict`, `residuals`, `ranef`);
4. parameter recovery under known DGP settings.

Run locally:

```{r run-tests, eval = FALSE}
devtools::test(filter = "brsmm")
```

## References

Ferrari, S. L. P. and Cribari-Neto, F. (2004). Beta regression for modelling
rates and proportions. *Journal of Applied Statistics*, 31(7), 799-815.
DOI: 10.1080/0266476042000214501. Validated online via:
<https://doi.org/10.1080/0266476042000214501> and
<https://econpapers.repec.org/RePEc:taf:japsta:v:31:y:2004:i:7:p:799-815>.

Lopes, J. E. (2024). *Beta Regression for Interval-Censored Scale-Derived
Outcomes*. MSc Dissertation, PPGMNE/UFPR.

Pinheiro, J. C. and Bates, D. M. (2000). *Mixed-Effects Models in S and S-PLUS*.
Springer. DOI: 10.1007/b98882. Validated online via:
<https://doi.org/10.1007/b98882> and
<https://books.google.com/books?id=N3WeyHFbHLQC>.
