---
title: "Analyst Tools for betaregscale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyst Tools for betaregscale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  digits = 4
)
```

## Overview

This vignette presents analyst-oriented tools added to **betaregscale**
for post-estimation workflows:

- `brs_table()`: model comparison tables.
- `brs_marginaleffects()`: average marginal effects (AME).
- `brs_predict_scoreprob()`: probabilities on the original score scale.
- `autoplot.brs()`: `ggplot2` diagnostics.
- `brs_cv()`: repeated k-fold cross-validation.

The examples use bounded scale data under the interval-censored beta
regression framework of Lopes (2024).

```{r library}
library(betaregscale)
```

## Mathematical foundations

### Complete likelihood by censoring type

For each observation \(i\), let \(\delta_i \in \{0,1,2,3\}\) indicate
exact, left-censored, right-censored, or interval-censored status.
With beta CDF \(F(\cdot)\), beta density \(f(\cdot)\), and interval
endpoints \([l_i, u_i]\), the contribution is:

\[
L_i(\theta)=
\begin{cases}
f(y_i; a_i, b_i), & \delta_i = 0,\\
F(u_i; a_i, b_i), & \delta_i = 1,\\
1 - F(l_i; a_i, b_i), & \delta_i = 2,\\
F(u_i; a_i, b_i)-F(l_i; a_i, b_i), & \delta_i = 3.
\end{cases}
\]

This is the basis for fitting, prediction, and validation metrics.

### Model-comparison metrics

`brs_table()` reports:

- \(\log L(\hat\theta)\),
- \(AIC = -2\log L(\hat\theta) + 2k\),
- \(BIC = -2\log L(\hat\theta) + k\log n\),

where \(k\) is the number of estimated parameters and \(n\) is sample size.

### Average marginal effects (AME)

`brs_marginaleffects()` computes AME by finite differences:

\[
\mathrm{AME}_j = \frac{1}{n}\sum_{i=1}^n
\frac{\hat g_i(x_{ij}+h)-\hat g_i(x_{ij})}{h},
\]

with \(\hat g_i\) on the requested prediction scale (`response` or `link`).
For binary covariates \(x_j \in \{0,1\}\), it uses the discrete contrast
\(\hat g(x_j=1)-\hat g(x_j=0)\).

### Score-scale probabilities

For integer scores \(s \in \{0,\dots,K\}\), `brs_predict_scoreprob()`
computes:

\[
P(Y=s)=
\begin{cases}
F(\mathrm{lim}/K), & s=0,\\
1-F((K-\mathrm{lim})/K), & s=K,\\
F((s+\mathrm{lim})/K)-F((s-\mathrm{lim})/K), & 1 \le s \le K-1.
\end{cases}
\]

These probabilities are directly aligned with interval geometry on the
original instrument scale.

### Cross-validation log score

In `brs_cv()`, fold-level predictive quality includes:

\[
\mathrm{log\_score} = \frac{1}{n_{test}}\sum_{i \in test}\log(p_i),
\]

where \(p_i\) is the predictive contribution from the same censoring-rule
piecewise definition shown above.

It also reports:

\[
\mathrm{RMSE}_{yt} = \sqrt{\frac{1}{n_{test}}\sum (y_{t,i}-\hat\mu_i)^2},
\qquad
\mathrm{MAE}_{yt} = \frac{1}{n_{test}}\sum |y_{t,i}-\hat\mu_i|.
\]

## Reproducible workflow

### 1) Simulate data and fit candidate models

```{r simulate-fit}
set.seed(2026)
n <- 220
d <- data.frame(
  x1 = rnorm(n),
  x2 = rnorm(n),
  z1 = rnorm(n)
)

sim <- brs_sim(
  formula = ~ x1 + x2 | z1,
  data = d,
  beta = c(0.20, -0.45, 0.25),
  zeta = c(0.15, -0.30),
  ncuts = 100,
  repar = 2
)

fit_fixed <- brs(y ~ x1 + x2, data = sim, repar = 2)
fit_var <- brs(y ~ x1 + x2 | z1, data = sim, repar = 2)
```

### 2) Compare models in one table

```{r table}
tab <- brs_table(
  fixed = fit_fixed,
  variable = fit_var,
  sort_by = "AIC"
)
knitr::kable(tab, digits = 4)
```

### 3) Estimate average marginal effects

```{r me}
me_mean <- brs_marginaleffects(
  fit_var,
  model = "mean",
  type = "response",
  interval = TRUE,
  n_sim = 120
)
knitr::kable(me_mean, digits = 4)

me_precision <- brs_marginaleffects(
  fit_var,
  model = "precision",
  type = "link",
  interval = TRUE,
  n_sim = 120,
  seed = 2026
)
knitr::kable(me_precision, digits = 4)
```

### 4) Predict score probabilities

```{r scoreprob}
P <- brs_predict_scoreprob(fit_var, scores = 0:10)
dim(P)
knitr::kable(P[1:6, 1:5], digits = 4)
knitr::kable(
  data.frame(row_sum = rowSums(P)),
  digits = 4
)
```

`rowSums(P)` should be close to 1 (up to numerical tolerance and score
subset selection).

### 5) ggplot2 diagnostics

```{r autoplot, eval = requireNamespace("ggplot2", quietly = TRUE)}
autoplot.brs(fit_var, type = "calibration")
autoplot.brs(fit_var, type = "score_dist", scores = 0:20)
autoplot.brs(fit_var, type = "cdf", max_curves = 4)
autoplot.brs(fit_var, type = "residuals_by_delta", residual_type = "rqr")
```

### 6) Repeated k-fold cross-validation

```{r cv}
cv_res <- brs_cv(
  y ~ x1 + x2 | z1,
  data = sim,
  k = 3,
  repeats = 1,
  repar = 2,
  seed = 2026
)

knitr::kable(cv_res, digits = 4)
knitr::kable(
  data.frame(
    metric = c("log_score", "rmse_yt", "mae_yt"),
    mean = colMeans(cv_res[, c("log_score", "rmse_yt", "mae_yt")], na.rm = TRUE)
  ),
  digits = 4
)
```

## Practical interpretation

- Prefer the model with lower AIC/BIC and better predictive `log_score`.
- Use AME on the response scale to communicate expected change in mean score
  (on the unit interval) from small covariate shifts.
- Use score probabilities to translate model outputs back to clinically
  interpretable scale categories.
- Inspect calibration and residual-by-censoring plots before inference.

## References

- Lopes, J. E. (2024). *Beta Regression for Interval-Censored
  Scale-Derived Outcomes*. MSc Dissertation, PPGMNE/UFPR.
- Ferrari, S. L. P., and Cribari-Neto, F. (2004). Beta regression for
  modelling rates and proportions. *Journal of Applied Statistics*,
  31(7), 799-815. DOI: 10.1080/0266476042000214501.
  Validated online via:
  <https://doi.org/10.1080/0266476042000214501> and
  <https://econpapers.repec.org/RePEc:taf:japsta:v:31:y:2004:i:7:p:799-815>.
- Hawker, G. A., Mian, S., Kendzerska, T., and French, M. (2011).
  Measures of adult pain: VAS, NRS, MPQ, SF-MPQ, CPGS, SF-36 BPS,
  and ICOAP. *Arthritis Care and Research*, 63(S11), S240-S252.
  DOI: 10.1002/acr.20543. Validated online via:
  <https://doi.org/10.1002/acr.20543> and
  <https://acrjournals.onlinelibrary.wiley.com/doi/10.1002/acr.20543>.
- Hjermstad, M. J., Fayers, P. M., Haugen, D. F., et al. (2011).
  Comparing NRS, VRS, and VAS pain scales in adults:
  a systematic review. *Journal of Pain and Symptom Management*,
  41(6), 1073-1093. DOI: 10.1016/j.jpainsymman.2010.08.016.
  Validated online via:
  <https://doi.org/10.1016/j.jpainsymman.2010.08.016> and
  <https://pubmed.ncbi.nlm.nih.gov/21621130/>.
