---
title: "Advanced Workflows for High-Level Users"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Workflows for High-Level Users}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(betaregscale)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  digits = 4
)
set.seed(2026)

kbl10 <- function(x, n = 10, digits = 4, ...) {
  knitr::kable(utils::head(as.data.frame(x), n), digits = digits, align = "c", ...)
}
```

## Audience and scope

This vignette is designed for analysts who are already familiar with beta regression and require a production-oriented workflow featuring:

1. reproducible model selection;
2. inferential and predictive diagnostics;
3. mixed-effects escalation (`brs` -> `brsmm`) with Likelihood Ratio (LR) comparisons;
4. high-signal reporting tables for technical decision-making.

```{r library}
library(betaregscale)
```

## 1) Reproducible simulation and data checks

```{r data-sim}
n <- 150
d <- data.frame(
  x1 = rnorm(n),
  x2 = rnorm(n),
  z1 = rnorm(n)
)

sim <- brs_sim(
  formula = ~ x1 + x2 | z1,
  data = d,
  beta = c(0.15, 0.55, -0.30),
  zeta = c(-0.10, 0.35),
  link = "logit",
  link_phi = "logit",
  ncuts = 100,
  repar = 2
)

kbl10(head(sim, 10))
kbl10(
  data.frame(
    n = nrow(sim),
    exact = sum(sim$delta == 0),
    left = sum(sim$delta == 1),
    right = sum(sim$delta == 2),
    interval = sum(sim$delta == 3)
  ),
  digits = 4
)
```

`brs_sim()` coarsens each draw exactly as the likelihood assumes, so the
censoring is non-informative. Its `delta` argument can force one type on
every row; `delta = 1` or `2` censors each row at the cell of its own
score, which is informative censoring with no finite MLE, and
`brs_sim()` warns. Leave `delta = NULL` for studies of the estimator.

## 2) Fixed-effects candidate set and model ranking

```{r fit-candidates}
fit_logit <- brs(y ~ x1 + x2 | z1, data = sim, link = "logit", repar = 2)
fit_probit <- brs(y ~ x1 + x2 | z1, data = sim, link = "probit", repar = 2)
fit_cauchit <- brs(y ~ x1 + x2 | z1, data = sim, link = "cauchit", repar = 2)

tab_rank <- brs_table(
  logit = fit_logit,
  probit = fit_probit,
  cauchit = fit_cauchit
)
kbl10(tab_rank)
```

## 3) Inference stack: Wald + bootstrap + AME



```{r inference-stack}
kbl10(brs_est(fit_logit))
kbl10(confint(fit_logit))

# Draws are kept so that the plots below can reuse this single fit rather
# than resampling a second time.
boot_tab <- brs_bootstrap(fit_logit, R = 30, level = 0.95, keep_draws = TRUE)
kbl10(head(boot_tab, 10))

set.seed(2026) # For marginal effects simulation
ame_mu <- brs_marginaleffects(
  fit_logit,
  model = "mean",
  type = "response",
  interval = TRUE,
  n_sim = 60,
  keep_draws = TRUE
)
kbl10(ame_mu)
```

Each bootstrap replicate draws a new response at the fitted shapes and
refits the original formula on a copy of the data, so factors,
transformations and prepared data keep their structure; each row keeps its
observation mechanism (exact, score, or analyst thresholds, which are treated
as fixed and non-informative). Refits start from the fitted estimate and use
the compiled Hessian. Failed replicates are counted in the printed header. `ci_type = "bca"` is also
available: with parametric resampling and a leave-one-out jackknife it is an
approximation (a warning says so once per session) and costs `R + n` fits
instead of `R`; see `?brs_bootstrap`.

```{r inference-visual}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  autoplot.brs_bootstrap(
    boot_tab,
    type = "ci_forest",
    title = "Bootstrap (percentile) vs Wald intervals"
  )
  autoplot.brs_marginaleffects(ame_mu, type = "forest")
}
```

## 4) Prediction layer on analyst scale

```{r prediction-layer}
score_prob <- brs_predict_scoreprob(fit_logit, scores = 0:10)
kbl10(score_prob[1:8, 1:7])
```

## 5) Out-of-sample validation



```{r cv-layer}
set.seed(2026) # For cross-validation reproducibility
cv_tab <- brs_cv(
  y ~ x1 + x2 | z1,
  data = sim,
  k = 5,
  repeats = 2,
  repar = 2
)
kbl10(head(cv_tab, 10))
```

## 6) Escalation to mixed models

### 6.1 Simulate clustered data with random intercept + slope

```{r mixed-data}
g <- 10
ni <- 25
id <- factor(rep(seq_len(g), each = ni))
n_mm <- length(id)
x1 <- rnorm(n_mm)
x2 <- rbinom(n_mm, size = 1, prob = 1 / 2)

b0 <- rnorm(g, sd = 0.40)
b1 <- rnorm(g, sd = 0.22)

eta_mu <- 0.20 + 0.65 * x1 - 0.30 * x2 + b0[id] + b1[id] * x1
eta_phi <- rep(-0.20, n_mm)

mu <- plogis(eta_mu)
phi <- plogis(eta_phi)
shp <- brs_repar(mu = mu, phi = phi, repar = 2)
y <- round(stats::rbeta(n_mm, shp$shape1, shp$shape2) * 100)

dmm <- data.frame(y = y, id = id, x1 = x1, x2 = x2)
kbl10(head(dmm, 10))
```

### 6.2 Fit evolutionary sequence

```{r mixed-fit}
fit_brs <- brs(y ~ x1 + x2, data = dmm, repar = 2)
fit_ri <- brsmm(y ~ x1 + x2, random = ~ 1 | id, data = dmm, repar = 2)
fit_rs <- brsmm(y ~ x1 + x2, random = ~ 1 + x1 | id, data = dmm, repar = 2)

tab_lr <- anova(fit_brs, fit_ri, fit_rs, test = "Chisq")
tab_lr
```

### 6.3 Model choice by LLR/LRT (ANOVA)

The `anova()` methods provide a practical Likelihood Ratio Test (LLR) workflow:

* `M0 = brs` (no random effects);
* `M1 = brsmm` with random intercept (`~ 1 | id`);
* `M2 = brsmm` with random intercept + slope (`~ 1 + x1 | id`).



In nested comparisons, the test statistic is:
$$LR=2\{\ell(\hat\theta_{\text{complex}})-\ell(\hat\theta_{\text{simple}})\}$$

Both steps add one random-effect term whose variance is on the boundary of
the parameter space under $H_0$ ($\sigma^2=0$). The reference distribution
is then the mixture $\frac12\chi^2_{d-1}+\frac12\chi^2_{d}$ (Self and
Liang, 1987; Stram and Lee, 1994): $\frac12\chi^2_0+\frac12\chi^2_1$ for
`M0 -> M1` and $\frac12\chi^2_1+\frac12\chi^2_2$ for `M1 -> M2`, where
the slope adds a variance and a covariance. `anova()` uses it for such rows
and says so in its heading; the naive $\chi^2_d$ p-value would be too
large. A term that adds almost nothing to the likelihood is also flagged by
`brsmm()` ("Variance component at the boundary").

```{r llr-decision}
tab_lr_df <- data.frame(model = rownames(tab_lr), tab_lr, row.names = NULL)
tab_lr_df$decision <- c(
  "baseline",
  ifelse(is.na(tab_lr_df$`Pr(>Chisq)`[2]), "inspect AIC/BIC + diagnostics",
    ifelse(tab_lr_df$`Pr(>Chisq)`[2] < 0.05, "prefer M1 over M0", "prefer M0 (parsimony)")
  ),
  ifelse(is.na(tab_lr_df$`Pr(>Chisq)`[3]), "inspect AIC/BIC + diagnostics",
    ifelse(tab_lr_df$`Pr(>Chisq)`[3] < 0.05, "prefer M2 over M1", "prefer M1 (parsimony)")
  )
)
kbl10(tab_lr_df)
```

## 7) Random-effects study (numeric + visual)

```{r ranef-study}
rs <- brsmm_re_study(fit_rs)
print(rs)
kbl10(rs$summary)
kbl10(rs$D)
kbl10(rs$Corr)
```



```{r ranef-plots}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  autoplot.brsmm(fit_rs, type = "ranef_caterpillar")
  autoplot.brsmm(fit_rs, type = "ranef_density")
  autoplot.brsmm(fit_rs, type = "ranef_pairs")
  autoplot.brsmm(fit_rs, type = "ranef_qq")
}
```

## 8) A Monte Carlo study (design of Lopes, 2023, ch. 3)

A simulation study is a loop of `brs_sim()` and `brs()` over a fixed
design. The code below mirrors the dissertation's design with a smaller
grid: repar 2 with logit links, `interval = "mid"`, two scale lengths
($K=10$ and $K=100$), two sample sizes and three dispersion levels,
200 replicates per cell. It takes under a minute, so it is not run here.

```{r mc-design, eval = FALSE}
R <- 200L
beta <- c(0.5, 0.2, -0.7)
grid <- expand.grid(sigma = c(0.08, 0.2, 0.4), n = c(100L, 500L), K = c(10L, 100L))
one_cell <- function(sigma, n, K) {
  set.seed(1000 + n + K + round(100 * sigma))
  d <- data.frame(x1 = rnorm(n), x2 = rbinom(n, 1, 0.5))   # fixed design
  truth <- c(beta, qlogis(sigma))
  est <- se <- matrix(NA_real_, R, 4)
  for (r in seq_len(R)) {
    s <- brs_sim(~ x1 + x2, data = d, beta = beta, phi = qlogis(sigma), ncuts = K)
    f <- tryCatch(brs(y ~ x1 + x2, data = s), error = function(e) NULL)
    if (is.null(f) || f$convergence != 0L) next
    est[r, ] <- coef(f)
    se[r, ] <- sqrt(diag(vcov(f)))
  }
  ok <- stats::complete.cases(est, se)
  err <- est[ok, ] - rep(truth, each = sum(ok))
  data.frame(sigma, n, K, par = c("b0", "b1", "b2", "logit_sigma"),
             bias = colMeans(err), rmse = sqrt(colMeans(err^2)),
             cover = colMeans(abs(err) <= qnorm(0.975) * se[ok, ]))
}
res <- do.call(rbind, Map(one_cell, grid$sigma, grid$n, grid$K))
```

Results for the slope $\beta_1=0.2$ and for $\mathrm{logit}(\sigma)$
(bias, root mean squared error and coverage of the 95% Wald interval,
with the default compiled Hessian; `hessian_method = "numDeriv"` gives the
same coverages):

| $K$ | $n$ | $\sigma$ | $\beta_1$ bias | RMSE | cover | $\mathrm{logit}\,\sigma$ bias | RMSE | cover |
|--:|--:|--:|--:|--:|--:|--:|--:|--:|
|  10 | 100 | 0.08 | 0.003 | 0.057 | 0.925 | -0.047 | 0.160 | 0.920 |
|  10 | 100 | 0.20 | 0.009 | 0.092 | 0.925 | -0.038 | 0.146 | 0.945 |
|  10 | 100 | 0.40 | -0.011 | 0.127 | 0.945 | -0.024 | 0.141 | 0.945 |
|  10 | 500 | 0.08 | 0.000 | 0.029 | 0.920 | -0.013 | 0.067 | 0.930 |
|  10 | 500 | 0.20 | 0.005 | 0.039 | 0.975 | -0.004 | 0.061 | 0.945 |
|  10 | 500 | 0.40 | -0.003 | 0.060 | 0.945 | -0.013 | 0.059 | 0.945 |
| 100 | 100 | 0.08 | -0.003 | 0.055 | 0.955 | -0.047 | 0.132 | 0.965 |
| 100 | 100 | 0.20 | 0.006 | 0.097 | 0.930 | -0.043 | 0.138 | 0.935 |
| 100 | 100 | 0.40 | 0.014 | 0.113 | 0.940 | -0.046 | 0.131 | 0.935 |
| 100 | 500 | 0.08 | 0.002 | 0.031 | 0.920 | -0.012 | 0.064 | 0.930 |
| 100 | 500 | 0.20 | 0.001 | 0.040 | 0.945 | -0.003 | 0.059 | 0.955 |
| 100 | 500 | 0.40 | 0.008 | 0.063 | 0.930 | -0.004 | 0.057 | 0.940 |

With 200 replicates the Monte Carlo standard error of a coverage near 0.95
is about 0.015. The slopes are unbiased at both scale lengths; the
dispersion is slightly underestimated at $n=100$ (about $-0.04$ on the logit
scale, a third of its RMSE) and the bias vanishes at $n=500$. A coarse
scale ($K=10$) costs little precision relative to $K=100$.

## 9) Practical decision checklist

* Start with `brs` candidates (`link`, `repar`) and rank them using `brs_table()`.
* Add bootstrap analysis and Average Marginal Effects (AME) before escalating complexity.
* Use `brs_cv()` for out-of-sample stability checks.
* Escalate to `brsmm` only when LR/AIC/BIC metrics and diagnostics clearly support it.
* For random slopes, always inspect `brsmm_re_study()` and the associated random-effects plots.
* Read the one-line warnings of `brs()`/`brsmm()` (gradient, Hessian, clamps, variance at the boundary) before any Wald statement; `fit$diagnostics` keeps the numbers.

## References

Lopes, J. E. (2023). *Modelos de regressão beta para dados de escala*.
Master's dissertation, Universidade Federal do Paraná, Curitiba.
<https://hdl.handle.net/1884/86624>.

Self, S. G., and Liang, K.-Y. (1987). Asymptotic properties of maximum
likelihood estimators and likelihood ratio tests under nonstandard
conditions. *Journal of the American Statistical Association*, 82(398),
605-610.

Stram, D. O., and Lee, J. W. (1994). Variance components testing in the
longitudinal mixed effects model. *Biometrics*, 50(4), 1171-1177.

Ferrari, S. L. P. and Cribari-Neto, F. (2004). Beta regression for modelling
rates and proportions. *Journal of Applied Statistics*, 31(7), 799-815.
DOI: 10.1080/0266476042000214501. Validated online via:
<https://doi.org/10.1080/0266476042000214501>.

Smithson, M., and Verkuilen, J. (2006). A better lemon squeezer? Maximum-likelihood
regression with beta-distributed dependent variables. *Psychological Methods*,
11(1), 54-71. DOI: 10.1037/1082-989X.11.1.54. Validated online via:
<https://doi.org/10.1037/1082-989X.11.1.54>.

Rue, H., Martino, S., and Chopin, N. (2009). Approximate Bayesian inference
for latent Gaussian models by using integrated nested Laplace approximations.
*Journal of the Royal Statistical Society: Series B*, 71(2), 319-392.
DOI: 10.1111/j.1467-9868.2008.00700.x. Validated online via:
<https://doi.org/10.1111/j.1467-9868.2008.00700.x>.
