---
title: "Introduction to betaregscale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to betaregscale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  digits = 4
)

kbl10 <- function(x, n = 10, digits = 4, ...) {
  knitr::kable(utils::head(as.data.frame(x), n), digits = digits, align = "c", ...)
}
```

## Overview

The **betaregscale** package provides maximum-likelihood estimation of beta
regression models for responses derived from bounded rating scales.
Common examples include pain intensity scales (NRS-11, NRS-21, NRS-101),
Likert-type scales, product quality ratings, and any instrument whose
response can be mapped to the open interval $(0,1)$.



The key idea is that a discrete score recorded on a bounded scale carries
measurement uncertainty inherent to the instrument. For instance, a pain
score of $y=6$ on a 0–10 NRS is not an exact value but rather
represents a range: after rescaling to $(0,1)$, the observation is
treated as interval-censored in $[0.55,0.65]$. The package uses the
beta distribution to model such data, building a complete likelihood that
supports mixed censoring types within the same dataset (Lopes, 2023).

## Installation

```{r install, eval = FALSE}
# Development version from GitHub:
# install.packages("remotes")
remotes::install_github("evandeilton/betaregscale")
```

```{r library}
library(betaregscale)
```

## The score scale and `ncuts`

`ncuts` is $K$, the **maximum score**. Scores run over $0, 1, \ldots, K$,
so the scale has $K + 1$ categories: an NRS-11 pain scale (0 to 10) has
`ncuts = 10`, a 0 to 100 scale `ncuts = 100`. The lowest score is the
left border and the highest the right border of the scale.

A scale that starts at 1 must be shifted to start at 0. A Likert item
1 to 5 becomes 0 to 4, with `ncuts = 4`:

```{r likert}
likert <- c(1, 2, 3, 5, 4, 1)
kbl10(brs_check(likert - 1, ncuts = 4))
```

Without the shift, category 1 would be read as an interior score and the
lower border of the scale would be a category that nobody can choose:

```{r likert-wrong}
brs_check(likert, ncuts = 5)[, "delta"]
```

## Likelihood and censoring types

The complete likelihood supports four censoring
types, automatically classified by `brs_check()`:

| $\delta$ | Type | Likelihood contribution |
|:--------:|:-----|:-----------------------|
| 0 | Exact (uncensored) | $f(y_i;a_i,b_i)$ |
| 1 | Left-censored ($y=0$) | $F(u_i;a_i,b_i)$ |
| 2 | Right-censored ($y=K$) | $1-F(l_i;a_i,b_i)$ |
| 3 | Interval-censored | $F(u_i;a_i,b_i)-F(l_i;a_i,b_i)$ |

where $f(\cdot)$ and $F(\cdot)$ are the beta density and CDF,
$[l_i,u_i]$ are the interval endpoints, and $(a_i,b_i)$ are the beta
shape parameters derived from $\mu_i$ and $\phi_i$ via the chosen
reparameterization. The likelihood is the product

$$L(\beta,\gamma)=\prod_{i=1}^n f(y_i)^{I(\delta_i=0)}\,F(u_i)^{I(\delta_i=1)}
\,\{1-F(l_i)\}^{I(\delta_i=2)}\,\{F(u_i)-F(l_i)\}^{I(\delta_i=3)}.$$

This is the complete likelihood of Lopes (2023). The table of censoring
types in the dissertation swaps the labels $\delta=1$ and $\delta=2$;
the package follows its equation. The cells are fixed by the scale, not by
the observed value, so the censoring is non-informative.

## Interval construction

A score $s$ is mapped to a cell of $(0,1)$. The direction of the
uncertainty interval is set by `interval` (the cases $m$, $r$ and $l$ of
Lopes, 2023):

| `interval` | cell of score $s$ | latent score of $y^*\in(0,1)$ |
|:--|:--|:--|
| `"mid"` (default) | $[s-h,\,s+h]/K$ | $Ky^*$ |
| `"right"` | $[s,\,s+1]/(K+1)$ | $(K+1)y^*$ |
| `"left"` | $[s,\,s+1]/(K+1)$ | $(K+1)y^*-1$ |

$K$ is the maximum score (`ncuts`) and $h$ the half-width (`lim`,
default **0.5**, used by `"mid"` only). The $K+1$ cells of `"right"` and
`"left"` are equal and partition $[0,1]$; this normalisation is a package
choice (the dissertation divides by $K$). `"right"` and `"left"` give the
same fit and differ only in `predict(type = "score")`. The three directions
are different coarsening models, so `anova()` refuses to compare them.

The borders need no special transformation: score 0 is left-censored and
score $K$ right-censored, and endpoints are only clamped to
$[10^{-5},1-10^{-5}]$ as a numerical guard. The fit therefore does not
depend on the sample size, unlike the $\{y(n-1)/R+1/2\}/n$ transformation
of Smithson and Verkuilen (2006).

```{r check-response}
# Illustrate brs_check with a 0-10 NRS scale
y_example <- c(0, 3, 5, 7, 10)
cr <- brs_check(y_example, ncuts = 10)
kbl10(cr)
```

The `delta` column shows that $y=0$ is left-censored ($\delta=1$),
$y=10$ is right-censored ($\delta=2$), and all interior values are
interval-censored ($\delta=3$). The same scores with right-direction cells:

```{r check-response-right}
kbl10(brs_check(y_example, ncuts = 10, interval = "right"))
```

Values already in $(0,1)$ are exact observations ($\delta=0$), one
observation at a time. Input that mixes such values with scores
$\geq 1$ is ambiguous and gives a warning:

```{r check-mixed, warning = TRUE}
kbl10(brs_check(c(0.3, 5, 10), ncuts = 10))
```

## Data preparation with `brs_prep()`

In practice, analysts may want to supply their own censoring indicators
or interval endpoints rather than relying on the automatic classification
of `brs_check()`. The `brs_prep()` function provides a flexible,
validated bridge between raw analyst data and `brs()`.

It supports four input modes:

### Mode 1: Score only (automatic)

```{r bs-prepare-mode1}
# Equivalent to brs_check - delta inferred from y
d1 <- data.frame(y = c(0, 3, 5, 7, 10), x1 = rnorm(5))
kbl10(brs_prep(d1, ncuts = 10))
```

### Mode 2: Score + explicit censoring indicator

```{r bs-prepare-mode2}
# Analyst specifies delta directly
d2 <- data.frame(
  y     = c(50, 0, 99, 50),
  delta = c(0, 1, 2, 3),
  x1    = rnorm(4)
)
kbl10(brs_prep(d2, ncuts = 100))
```

### Mode 3: Interval endpoints with NA patterns

When the analyst provides `left` and/or `right` columns, censoring is
inferred from the NA pattern. Bounds are on the latent score scale and
must lie in $[-0.5, K+0.5]$ under `"mid"` ($[0,K+1]$ and $[-1,K]$ under
`"right"` and `"left"`). An interval that reaches 0 or 1 on the unit scale
becomes left- or right-censored; one that reaches both covers the whole
scale, carries no information and gives a warning.

```{r bs-prepare-mode3}
d3 <- data.frame(
  left  = c(NA, 20, 30, NA),
  right = c(5, NA, 45, NA),
  y     = c(NA, NA, NA, 50),
  x1    = rnorm(4)
)
kbl10(brs_prep(d3, ncuts = 100))
```

### Mode 4: Analyst-supplied intervals

When the analyst provides `y`, `left`, and `right` simultaneously, their
endpoints are used directly (rescaled by $K$ under `"mid"`):

```{r bs-prepare-mode4}
d4 <- data.frame(
  y     = c(50, 75),
  left  = c(48, 73),
  right = c(52, 77),
  x1    = rnorm(2)
)
kbl10(brs_prep(d4, ncuts = 100))
```

### Using brs_prep with `brs()`

Data processed by `brs_prep()` is automatically detected by
`brs()` - the internal `brs_check()` step is skipped:

```{r prepare-workflow}
set.seed(42)
n <- 200
dat <- data.frame(x1 = rnorm(n), x2 = rnorm(n))
sim <- brs_sim(
  formula = ~ x1 + x2, data = dat,
  beta = c(0.2, -0.5, 0.3), phi = 0.3,
  link = "logit", link_phi = "logit",
  repar = 2
)
# Remove left, right, yt so brs_prep can rebuild them
prep <- brs_prep(sim[-c(1:3)], ncuts = 100)
fit_prep <- brs(y ~ x1 + x2,
  data = prep, repar = 2,
  link = "logit", link_phi = "logit"
)
summary(fit_prep, digits = 4)
```

## Example 1: Fixed dispersion model

### Simulating data

We simulate observations from a beta regression model with fixed
dispersion, two covariates, and a logit link for the mean.

```{r sim-fixed}
set.seed(4255)
n <- 250
dat <- data.frame(x1 = rnorm(n), x2 = rnorm(n))

sim_fixed <- brs_sim(
  formula  = ~ x1 + x2,
  data     = dat,
  beta     = c(0.3, -0.6, 0.4),
  phi      = 1 / 10,
  link     = "logit",
  link_phi = "logit",
  ncuts    = 100,
  repar    = 2
)

kbl10(head(sim_fixed, 8))
```

Each observation is centered in its interval. For example, a score of 67
on a 0–100 scale yields $y_t=0.67$ with the interval $[0.665,0.675]$.

### Fitting the model

```{r fit-fixed}
fit_fixed <- brs(
  y ~ x1 + x2,
  data     = sim_fixed,
  link     = "logit",
  link_phi = "logit",
  repar    = 2
)
summary(fit_fixed)
```

The summary output follows the `betareg` package style, showing separate
coefficient tables for the mean and precision submodels, with Wald
z-tests and $p$-values based on the standard normal distribution.

### Estimation and fit diagnostics

`optim()` (BFGS) maximises the log-likelihood, evaluated in C++. The
gradient and, by default, the Hessian use the chain rule on the two linear
predictors, with per-observation derivatives by Richardson central
differences (`hessian_method = "cpp"`; `"numDeriv"` is also available).
Standard errors come from the inverse of the negative Hessian; no
generalised inverse is used. `brs()` also accepts `start` and `control`
(merged into `list(maxit = 5000)`).

After the fit, `brs()` checks the result and stores the numbers:

```{r diagnostics}
str(fit_fixed$diagnostics[c("grad_gain", "grad_step", "min_eig_scaled", "hessian_nd", "n_clamped")])
```

Each problem gives a one-line warning:

| Warning | Meaning | What to do |
|:--|:--|:--|
| model matrix is rank deficient (error) | aliased columns | remove the named columns |
| model matrix is nearly collinear | condition number above $10^4$ | drop, combine or centre the named columns |
| Gradient not ~0 at the estimate | a Newton step would still gain more than 0.01 in log-likelihood | refit with `method = "L-BFGS-B"`; rescale covariates whose scales differ widely (the message says so) |
| Hessian near-singular or not negative definite | a direction is nearly flat (not identified, or covariates on very different scales) | `vcov()` gives `NA` where it cannot estimate; rescale, use LR tests or the bootstrap, simplify |
| observations on the clamp boundary | means or shapes at $10^{-5}$ or $1-10^{-5}$, e.g. a group entirely at one border | merge sparse groups or drop the separating covariate |

The classic case of the last two is a data set where every observation is
censored on the same side: the likelihood has no finite maximum.

### Goodness of fit

```{r gof-fixed}
kbl10(brs_gof(fit_fixed))
```

### Comparing link functions

The package supports several link functions for the mean submodel. We
can compare them using information criteria:

```{r compare-links}
links <- c("logit", "probit", "cauchit", "cloglog")
fits <- lapply(setNames(links, links), function(lnk) {
  brs(y ~ x1 + x2, data = sim_fixed, link = lnk, repar = 2)
})

# Estimates
est_table <- do.call(rbind, lapply(names(fits), function(lnk) {
  e <- brs_est(fits[[lnk]])
  e$link <- lnk
  e
}))
kbl10(est_table)

# Goodness of fit
gof_table <- do.call(rbind, lapply(fits, brs_gof))
kbl10(gof_table)
```

### Residual diagnostics

`residuals()` offers several types. Randomized quantile residuals
(`type = "rqr"`, Dunn and Smyth, 1996) draw $u_i$ uniformly on
$(F(l_i),F(u_i))$ (or $(0,F(u_i))$, $(F(l_i),1)$ at the borders) and
return $\Phi^{-1}(u_i)$: they are standard normal under the model whatever
the censoring, and are the recommended type. Pearson, deviance and
weighted residuals are evaluated at the cell centre (the midpoint
convention of Lopes, 2023) and are unreliable at the borders of the scale.
`summary()` draws the quantile residuals without changing the RNG state.

The `plot()` method provides six diagnostic panels. By default, the
first four are shown:

```{r plot-fixed, fig.height = 6}
plot(fit_fixed, caption = NULL, gg = TRUE, title = NULL, theme = ggplot2::theme_bw())
```

For ggplot2 output (requires the **ggplot2** package):

```{r plot-fixed-gg, eval = requireNamespace("ggplot2", quietly = TRUE), fig.height = 6}
plot(fit_fixed, gg = TRUE)
```

### Predictions

```{r predict-fixed}
# Fitted means
kbl10(
  data.frame(mu_hat = head(predict(fit_fixed, type = "response"))),
  digits = 4
)

# Conditional variance
kbl10(
  data.frame(var_hat = head(predict(fit_fixed, type = "variance"))),
  digits = 4
)

# Quantile predictions
kbl10(predict(fit_fixed, type = "quantile", at = c(0.10, 0.25, 0.5, 0.75, 0.90)))
```

### Confidence intervals

Wald confidence intervals based on the asymptotic normal approximation:

```{r confint-fixed}
kbl10(confint(fit_fixed))
kbl10(confint(fit_fixed, model = "mean"))
```

### Censoring structure

The `brs_cens()` function provides a visual and tabular
overview of the censoring types in the fitted model:

```{r censoring-summary, fig.height = 5}
brs_cens(fit_fixed, gg = TRUE, inform = TRUE)
```

## Example 2: Variable dispersion model

In many applications, the dispersion parameter $\phi$ may depend on
covariates. The package supports variable-dispersion models using the
`Formula` package notation: `y ~ x1 + x2 | z1 + z2`, where the terms
after `|` define the linear predictor for $\phi$. The same `brs_sim()`
function is used for fixed and variable dispersion; the second formula
part activates the precision submodel in simulation.

### Simulating data

```{r sim-variable}
set.seed(2222)
n <- 250
dat_z <- data.frame(
  x1 = rnorm(n),
  x2 = rnorm(n),
  x3 = rbinom(n, size = 1, prob = 0.5),
  z1 = rnorm(n),
  z2 = rnorm(n)
)

sim_var <- brs_sim(
  formula = ~ x1 + x2 + x3 | z1 + z2,
  data = dat_z,
  beta = c(0.2, -0.6, 0.2, 0.2),
  zeta = c(0.2, -0.8, 0.6),
  link = "logit",
  link_phi = "logit",
  ncuts = 100,
  repar = 2
)

kbl10(head(sim_var, 10))
```

### Fitting the model

```{r fit-variable}
fit_var <- brs(
  y ~ x1 + x2 | z1,
  data     = sim_var,
  link     = "logit",
  link_phi = "logit",
  repar    = 2
)
summary(fit_var)
```

Notice the `(phi)_` prefix in the precision coefficient names, following
the `betareg` convention.

### Accessing coefficients by submodel

```{r coef-variable}
# Full parameter vector
round(coef(fit_var), 4)

# Mean submodel only
round(coef(fit_var, model = "mean"), 4)

# Precision submodel only
round(coef(fit_var, model = "precision"), 4)

# Variance-covariance matrix for the mean submodel
kbl10(vcov(fit_var, model = "mean"))
```

### Comparing link functions (variable dispersion)

```{r compare-links-var}
links <- c("logit", "probit", "cauchit", "cloglog")
fits_var <- lapply(setNames(links, links), function(lnk) {
  brs(y ~ x1 + x2 | z1, data = sim_var, link = lnk, repar = 2)
})

# Estimates
est_var <- do.call(rbind, lapply(names(fits_var), function(lnk) {
  e <- brs_est(fits_var[[lnk]])
  e$link <- lnk
  e
}))
kbl10(est_var)

# Goodness of fit
gof_var <- do.call(rbind, lapply(fits_var, brs_gof))
kbl10(gof_var)
```

### Diagnostics for variable dispersion

```{r plot-variable, fig.height = 6}
plot(fit_var)
```

## Advanced analyst functions

The package includes analyst-facing helpers for uncertainty quantification,
effect interpretation, score-scale communication, and predictive validation.

### Parametric bootstrap confidence intervals

Each replicate draws a new response from the fitted model and refits the
original formula on a copy of the data. Each row keeps its observation
mechanism: exact values stay continuous, scores are re-coarsened on the
fit's grid, and analyst thresholds from `brs_prep()` are kept as fixed,
non-informative thresholds (the censoring type is re-drawn by the side of
them where the new value falls). Refits start from the fitted estimate.
Failed replicates are counted and printed. Use `R` of at least 199 in
practice; `ci_type = "bca"` costs one extra refit per observation.

```{r bootstrap-fixed}
set.seed(101)
boot_ci <- brs_bootstrap(
  fit_fixed,
  R = 30,
  level = 0.95,
  ci_type = "percentile",
  keep_draws = TRUE
)
kbl10(head(boot_ci, 10))
```

```{r bootstrap-forest, eval = requireNamespace("ggplot2", quietly = TRUE)}
autoplot.brs_bootstrap(
  boot_ci,
  type = "ci_forest",
  title = "Bootstrap (percentile) vs Wald intervals"
)
```

### Average marginal effects (AME)



```{r marginaleffects-fixed}
set.seed(202)
ame <- brs_marginaleffects(
  fit_fixed,
  model = "mean",
  type = "response",
  interval = TRUE,
  n_sim = 60,
  keep_draws = TRUE
)
kbl10(ame)
```

```{r marginaleffects-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
autoplot.brs_marginaleffects(ame, type = "forest")
```

### Score probabilities on the original scale

```{r scoreprob-fixed}
prob_scores <- brs_predict_scoreprob(fit_fixed, scores = 0:10)
kbl10(prob_scores[1:6, 1:6])
kbl10(
  data.frame(row_sum = rowSums(prob_scores)[1:6]),
  digits = 4
)
```

### Repeated k-fold cross-validation

```{r cv-fixed}
set.seed(303) # For cross-validation reproducibility
cv_res <- brs_cv(
  y ~ x1 + x2,
  data = sim_fixed,
  k = 5,
  repeats = 2,
  repar = 2,
)
kbl10(cv_res)
kbl10(
  data.frame(
    metric = c("log_score", "rmse_yt", "mae_yt"),
    mean = c(
      mean(cv_res$log_score, na.rm = TRUE),
      mean(cv_res$rmse_yt, na.rm = TRUE),
      mean(cv_res$mae_yt, na.rm = TRUE)
    )
  ),
  digits = 4
)
```

## S3 methods reference

The following standard S3 methods are available for objects of class
`"brs"`:

| Method | Description |
|:-------|:-----------|
| `print()` | Compact display of call and coefficients |
| `summary()` | Detailed output with Wald tests and goodness-of-fit |
| `coef(model=)` | Extract coefficients (full, mean, or precision) |
| `vcov(model=)` | Variance-covariance matrix (full, mean, or precision) |
| `confint(model=)` | Wald confidence intervals |
| `logLik()` | Log-likelihood value |
| `AIC()`, `BIC()` | Information criteria |
| `nobs()` | Number of observations |
| `formula()` | Model formula |
| `model.matrix(model=)` | Design matrix (mean or precision) |
| `fitted()` | Fitted mean values |
| `residuals(type=)` | Residuals: response, pearson, deviance, rqr, weighted, sweighted |
| `predict(type=)` | Predictions: response, link, precision, variance, quantile, score, expected_score |
| `plot(gg=)` | Diagnostic plots (base R or ggplot2) |

## Reparameterizations

The package supports three reparameterizations of the beta distribution,
controlled by the `repar` argument. Each models different parameters, so
the admissible links differ; the first link of each cell is the default.

| `repar` | parameters | shapes $(a,b)$ | `link` | `link_phi` |
|:--|:--|:--|:--|:--|
| 0 | shapes $p,q>0$ | $(p,q)$ | log, sqrt | log, sqrt |
| 1 | mean $\mu$, precision $\phi>0$ | $(\mu\phi,(1-\mu)\phi)$ | logit, probit, cauchit, cloglog | log, sqrt |
| 2 | mean $\mu$, dispersion $\phi\in(0,1)$ | $(\mu(1-\phi)/\phi,(1-\mu)(1-\phi)/\phi)$ | logit, probit, cauchit, cloglog | logit, probit, cauchit, cloglog |

In every scheme $E[Y]=a/(a+b)$ and $\mathrm{Var}[Y]=E[Y](1-E[Y])/(1+a+b)$.
Under `repar = 1` this is $\mu(1-\mu)/(1+\phi)$ (Ferrari and
Cribari-Neto, 2004); under `repar = 2` it is $\phi\,\mu(1-\mu)$, so the
dispersion is $\phi=1/(1+a+b)$, the variance relative to its maximum
$\mu(1-\mu)$ (not a coefficient of variation). Under `repar = 0` the
first parameter is the shape $p$ and the mean is $p/(p+q)$.

```{r reparam}
# Precision parameterization: mu = 0.5, phi = 10 (high precision)
brs_repar(mu = 0.5, phi = 10, repar = 1)

# Mean-variance parameterization: mu = 0.5, phi = 0.1 (low dispersion)
brs_repar(mu = 0.5, phi = 0.1, repar = 2)
```

## References

- Lopes, J. E. (2023). *Modelos de regressão beta para dados de escala*.
  Master's dissertation, Universidade Federal do Paraná, Curitiba.
  <https://hdl.handle.net/1884/86624>.

- Dunn, P. K., and Smyth, G. K. (1996). Randomized quantile residuals.
  *Journal of Computational and Graphical Statistics*, **5**(3), 236–244.

- Ferrari, S. L. P., and Cribari-Neto, F. (2004). Beta regression for
  modelling rates and proportions. *Journal of Applied Statistics*,
  **31**(7), 799–815. DOI: 10.1080/0266476042000214501.
  Validated online via:
  <https://doi.org/10.1080/0266476042000214501>.

- Smithson, M., and Verkuilen, J. (2006). A better lemon squeezer?
  Maximum-likelihood regression with beta-distributed dependent variables.
  *Psychological Methods*, **11**(1), 54–71.
  DOI: 10.1037/1082-989X.11.1.54.
  Validated online via:
  <https://doi.org/10.1037/1082-989X.11.1.54>.

- Hawker, G. A., Mian, S., Kendzerska, T., and French, M. (2011).
  Measures of adult pain: Visual Analog Scale for Pain (VAS Pain),
  Numeric Rating Scale for Pain (NRS Pain), McGill Pain Questionnaire (MPQ),
  Short-Form McGill Pain Questionnaire (SF-MPQ), Chronic Pain Grade Scale
  (CPGS), Short Form-36 Bodily Pain Scale (SF-36 BPS), and Measure of
  Intermittent and Constant Osteoarthritis Pain (ICOAP).
  *Arthritis Care and Research*, **63**(S11), S240–S252.
  DOI: 10.1002/acr.20543.
  Validated online via:
  <https://doi.org/10.1002/acr.20543>.

- Hjermstad, M. J., Fayers, P. M., Haugen, D. F., et al. (2011).
  Studies comparing numerical rating scales, verbal rating scales, and
  visual analogue scales for assessment of pain intensity in adults:
  a systematic literature review.
  *Journal of Pain and Symptom Management*, **41**(6), 1073–1093.
  DOI: 10.1016/j.jpainsymman.2010.08.016.
  Validated online via:
  <https://doi.org/10.1016/j.jpainsymman.2010.08.016>.
  
