---
title: "Mixed-Effects Beta Interval Regression with brsmm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Effects Beta Interval Regression with brsmm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  digits = 4
)

kbl10 <- function(x, n = 10, digits = 4, ...) {
  knitr::kable(utils::head(as.data.frame(x), n), digits = digits, align = "c", ...)
}
```

## Overview

`brsmm()` extends `brs()` to clustered data by adding Gaussian random
effects in the mean submodel while preserving the interval-censored
beta likelihood for scale-derived outcomes.

This vignette covers:

1. full model mathematics;
2. estimation by marginal maximum likelihood (Laplace, adaptive quadrature or quasi-Monte Carlo);
3. practical use of all current `brsmm` methods;
4. inferential and validation workflows, including parameter recovery.

```{r library}
library(betaregscale)
```

## Mathematical model

Assume observations $i = 1, \dots, n_j$ within groups
$j = 1, \dots, G$, with group-specific random-effects vector
$\mathbf{b}_j \in \mathbb{R}^{q_b}$.

### Linear predictors

$$
\eta_{\mu,ij} = x_{ij}^\top \beta + w_{ij}^\top \mathbf{b}_j,
\qquad
\eta_{\phi,ij} = z_{ij}^\top \gamma
$$

$$
\mu_{ij} = g^{-1}(\eta_{\mu,ij}),
\qquad
\phi_{ij} = h^{-1}(\eta_{\phi,ij})
$$

with $g(\cdot)$ and $h(\cdot)$ chosen by `link` and `link_phi`.
The random-effects design row $w_{ij}$ is defined by `random = ~ terms | group`.

### Beta parameterization

For each $(\mu_{ij},\phi_{ij})$, `repar` maps to beta shape parameters
$(a_{ij},b_{ij})$ via `brs_repar()`.

### Conditional contribution by censoring type

Each observation contributes:

$$
L_{ij}(b_j;\theta)=
\begin{cases}
f(y_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=0\\
F(u_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=1\\
1 - F(l_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=2\\
F(u_{ij}; a_{ij}, b_{ij}) - F(l_{ij}; a_{ij}, b_{ij}), & \delta_{ij}=3
\end{cases}
$$

where $l_{ij},u_{ij}$ are interval endpoints on $(0,1)$,
$f(\cdot)$ is beta density, and $F(\cdot)$ is beta CDF.

### Random-effects distribution

$$
\mathbf{b}_j \sim \mathcal{N}(\mathbf{0}, D),
$$

where $D$ is a symmetric positive-definite covariance matrix.
Internally, `brsmm()` optimizes a packed lower-Cholesky parameterization
$D = LL^\top$ (diagonal entries on log-scale for positivity).

### Group marginal likelihood

$$
L_j(\theta)=\int_{\mathbb{R}^{q_b}}
\prod_{i=1}^{n_j} L_{ij}(b_j;\theta)\,
\varphi_{q_b}(\mathbf{b}_j;\mathbf{0},D)\,d\mathbf{b}_j
$$

$$
\ell(\theta)=\sum_{j=1}^G \log L_j(\theta)
$$

### Laplace approximation used by `brsmm()`

Define
$$
Q_j(\mathbf{b})=
\sum_{i=1}^{n_j}\log L_{ij}(\mathbf{b};\theta)+
\log\varphi_{q_b}(\mathbf{b};\mathbf{0},D)
$$
and $\hat{\mathbf{b}}_j=\arg\max_{\mathbf{b}} Q_j(\mathbf{b})$, with curvature
$$
H_j = -\nabla^2 Q_j(\hat{\mathbf{b}}_j).
$$
Then

$$
\log L_j(\theta) \approx
Q_j(\hat{\mathbf{b}}_j) +
\frac{q_b}{2}\log(2\pi) -
\frac{1}{2}\log|H_j|.
$$

`brsmm()` maximizes the approximated $\ell(\theta)$ with
`stats::optim()`, and computes group-level posterior modes
$\hat{\mathbf{b}}_j$. For $q_b = 1$, this reduces to the scalar
random-intercept formula.

### Other integration methods and estimation details

* `int_method = "aghq"`: adaptive Gauss-Hermite quadrature with
  `n_points` nodes per dimension, placed at
  $\hat{\mathbf{b}}_j+\sqrt{2}\,H_j^{-1/2}\mathbf{z}$ with the symmetric
  square root $H_j^{-1/2}$ (`n_points`$^{q_b}$ nodes in total, at most 500000).
* `int_method = "qmc"`: importance sampling from
  $\mathcal{N}(\hat{\mathbf{b}}_j, H_j^{-1})$ with `qmc_points` Halton points.
  It is deterministic and, with two or more random effects, underestimates
  the log-likelihood (about $-0.05$ at 1024 points); prefer `"aghq"` for up
  to three random effects.

The inner mode is found by a Levenberg-Marquardt Newton method, warm-started
from the previous evaluation (the cache is cleared at the start of each fit).
A group whose curvature is not positive definite at its mode adds the
penalty value, and `brsmm()` warns when that happens at the estimate.
`optim()` receives the compiled gradient of the chosen approximation (chain
rule on the linear predictor and implicit-function theorem at the modes),
and the Hessian behind `vcov()` is a Richardson central difference of that
gradient (`hessian_method = "cpp"`).

## Simulating clustered scale data

The next helper simulates data from a known mixed model to illustrate fitting,
inference, and recovery checks.

```{r simulate-data}
sim_brsmm_data <- function(seed = 3501L, g = 24L, ni = 12L,
                           beta = c(0.20, 0.65),
                           gamma = c(-0.15),
                           sigma_b = 0.55) {
  set.seed(seed)
  id <- factor(rep(seq_len(g), each = ni))
  n <- length(id)
  x1 <- rnorm(n)

  b_true <- rnorm(g, mean = 0, sd = sigma_b)
  eta_mu <- beta[1] + beta[2] * x1 + b_true[as.integer(id)]
  eta_phi <- rep(gamma[1], n)

  mu <- plogis(eta_mu)
  phi <- plogis(eta_phi)
  shp <- brs_repar(mu = mu, phi = phi, repar = 2)
  y <- round(stats::rbeta(n, shp$shape1, shp$shape2) * 100)

  list(
    data = data.frame(y = y, x1 = x1, id = id),
    truth = list(beta = beta, gamma = gamma, sigma_b = sigma_b, b = b_true)
  )
}

sim <- sim_brsmm_data(
  g = 12,
  ni = 20,
  beta = c(0.20, 0.65),
  gamma = c(-0.15),
  sigma_b = 0.55
)
str(sim$data)
```

## Fitting `brsmm()`

```{r fit-brsmm}
fit_mm <- brsmm(
  y ~ x1,
  random = ~ 1 | id,
  data = sim$data,
  repar = 2,
  int_method = "laplace",
  method = "BFGS",
  control = list(maxit = 1000)
)

summary(fit_mm)
```

## Random intercept + slope example

The model below includes a random intercept and random slope for `x1`:

```{r fit-brsmm-random-slope}
fit_mm_rs <- brsmm(
  y ~ x1,
  random = ~ 1 + x1 | id,
  data = sim$data,
  repar = 2,
  int_method = "laplace",
  method = "BFGS",
  control = list(maxit = 1200)
)

summary(fit_mm_rs)
```

The random-effects block of `summary()` reports standard deviations and the
correlation, with 95% intervals built on the log scale (SD) and the
$\mathrm{atanh}$ scale (correlation) and mapped back; it gives no test. The
same numbers are in `summary(fit_mm_rs)$varcorr`.

Covariance structure of random effects:

```{r random-covariance}
kbl10(fit_mm_rs$random$D)
kbl10(
  data.frame(term = names(fit_mm_rs$random$sd_b), sd = as.numeric(fit_mm_rs$random$sd_b)),
  digits = 4
)
kbl10(head(ranef(fit_mm_rs), 10))
```

## Additional studies of random effects (numerical and visual)

Following practices from established mixed-models packages, the package now allows for a dedicated study of the random effects focusing on:

* $D$ structure and correlation;
* empirical distribution of modes by group;
* empirical shrinkage intensity;
* specific visual diagnostics for the random components.

The intraclass correlation reported by `brsmm_re_study()` is the
correlation of $\mathrm{logit}(Y)$ between two observations of the same
group with the same covariates. With $m(b)=\psi(a)-\psi(b)$ and
$v(b)=\psi_1(a)+\psi_1(b)$ (mean and variance of $\mathrm{logit}\,Y$ for
$Y\sim\mathrm{Beta}(a,b)$, with the shapes at random effect $b$),
$$
\mathrm{ICC}=\frac{\mathrm{Var}_b[m(b)]}{\mathrm{Var}_b[m(b)]+E_b[v(b)]},
$$
computed by Gauss-Hermite quadrature over $b\sim N(0,\,w^\top D w)$ and
averaged over the observations. It uses the beta level-1 variance: the
logistic value $\pi^2/3$ of binary models does not describe a beta
response. The moments of $\mathrm{logit}(Y)$ can be infinite (probit link
with $\sigma_b^2\geq 1/2$, cloglog link, or a very large $\sigma_b$); the
ICC is then `NA` with a warning, because its value would be set by the
numerical clamp of the mean.

```{r ranef-study}
re_study <- brsmm_re_study(fit_mm_rs)
print(re_study)
kbl10(re_study$summary)
kbl10(re_study$D)
kbl10(re_study$Corr)
```

Suggested visualizations for random effects:

```{r ranef-visuals}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  autoplot.brsmm(fit_mm_rs, type = "ranef_caterpillar")
  autoplot.brsmm(fit_mm_rs, type = "ranef_density")
  autoplot.brsmm(fit_mm_rs, type = "ranef_pairs")
  autoplot.brsmm(fit_mm_rs, type = "ranef_qq")
}
```

## Core methods

### Coefficients and random effects

`coef(fit_mm, model = "random")` returns packed random-effect covariance
parameters on the optimizer scale (lower-Cholesky, with a log-diagonal).
For random-intercept models, this simplifies to $\log \sigma_b$.

```{r methods-coef}
kbl10(
  data.frame(
    parameter = names(coef(fit_mm, model = "full")),
    estimate = as.numeric(coef(fit_mm, model = "full"))
  ),
  digits = 4
)
kbl10(
  data.frame(
    log_sigma_b = as.numeric(coef(fit_mm, model = "random")),
    sigma_b = as.numeric(exp(coef(fit_mm, model = "random")))
  ),
  digits = 4
)
kbl10(head(ranef(fit_mm), 10))
```

For random intercept + slope models:

```{r methods-coef-rs}
kbl10(
  data.frame(
    parameter = names(coef(fit_mm_rs, model = "random")),
    estimate = as.numeric(coef(fit_mm_rs, model = "random"))
  ),
  digits = 4
)
kbl10(fit_mm_rs$random$D)
```

### Variance-covariance, summary and likelihood criteria

```{r methods-summary}
vc <- vcov(fit_mm)
dim(vc)

sm <- summary(fit_mm)
kbl10(sm$coefficients$mean)
kbl10(sm$varcorr)

kbl10(
  data.frame(
    logLik = as.numeric(logLik(fit_mm)),
    AIC = AIC(fit_mm),
    BIC = BIC(fit_mm),
    nobs = nobs(fit_mm)
  ),
  digits = 4
)
```

### Fitted values, prediction and residuals

```{r methods-predict}
kbl10(
  data.frame(
    mu_hat = head(fitted(fit_mm, type = "mu")),
    phi_hat = head(fitted(fit_mm, type = "phi")),
    pred_mu = head(predict(fit_mm, type = "response")),
    pred_eta = head(predict(fit_mm, type = "link")),
    pred_phi = head(predict(fit_mm, type = "precision")),
    pred_var = head(predict(fit_mm, type = "variance"))
  ),
  digits = 4
)

kbl10(
  data.frame(
    res_response = head(residuals(fit_mm, type = "response")),
    res_pearson = head(residuals(fit_mm, type = "pearson"))
  ),
  digits = 4
)
```

### Diagnostic plotting methods

`plot.brsmm()` supports base and ggplot2 backends:

```{r methods-plot}
plot(fit_mm, which = 1:4, type = "pearson")

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot(fit_mm, which = 1:2, gg = TRUE)
}
```

`autoplot.brsmm()` provides focused ggplot diagnostics:

```{r methods-autoplot}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  autoplot.brsmm(fit_mm, type = "calibration")
  autoplot.brsmm(fit_mm, type = "score_dist")
  autoplot.brsmm(fit_mm, type = "ranef_qq")
  autoplot.brsmm(fit_mm, type = "residuals_by_group")
}
```

### Prediction with `newdata`

If `newdata` contains unseen groups, `predict.brsmm()` uses a random effect
equal to zero for those levels.

```{r methods-newdata}
nd <- sim$data[1:8, c("x1", "id")]
kbl10(
  data.frame(pred_seen = as.numeric(predict(fit_mm, newdata = nd, type = "response"))),
  digits = 4
)

nd_unseen <- nd
nd_unseen$id <- factor(rep("new_cluster", nrow(nd_unseen)))
kbl10(
  data.frame(pred_unseen = as.numeric(predict(fit_mm, newdata = nd_unseen, type = "response"))),
  digits = 4
)
```

The same logic applies to random intercept + slope models:

```{r methods-newdata-rs}
kbl10(
  data.frame(pred_rs_seen = as.numeric(predict(fit_mm_rs, newdata = nd, type = "response"))),
  digits = 4
)
kbl10(
  data.frame(pred_rs_unseen = as.numeric(predict(fit_mm_rs, newdata = nd_unseen, type = "response"))),
  digits = 4
)
```

## Statistical tests and validation workflow

### Wald tests (from `summary`)

`summary.brsmm()` reports Wald $z$-tests for the fixed effects:
$$
z_k = \hat\theta_k / \mathrm{SE}(\hat\theta_k).
$$
It gives no test for the random effects: a $z$-test of $\log\sigma_b$
tests $\sigma_b=1$, and $\sigma_b=0$ lies on the boundary. Their SD and
correlation come with intervals only; test a variance component with the
likelihood-ratio test below.

```{r wald-tests}
sm <- summary(fit_mm)
kbl10(sm$coefficients$mean)
kbl10(sm$coefficients$precision)
```

### Evolutionary scheme and Likelihood Ratio (LR) test selection

A practical workflow of increasing complexity:

1.  `brs()`: no random effect (ignores clustering);
2.  `brsmm(..., random = ~ 1 | id)`: random intercept;
3.  `brsmm(..., random = ~ 1 + x1 | id)`: random intercept + slope.

In both jumps the added variance lies on the boundary of the parameter
space under $H_0$, so the likelihood ratio does not follow $\chi^2_d$. With
one added random-effect term the reference is the mixture
$\frac12\chi^2_{d-1}+\frac12\chi^2_d$: $\frac12\chi^2_0+\frac12\chi^2_1$
from `brs` to the random intercept, $\frac12\chi^2_1+\frac12\chi^2_2$ from
the intercept to intercept + slope (Self and Liang, 1987; Stram and Lee,
1994). `anova()` uses it for these rows and explains it in its heading; the
naive $\chi^2_d$ p-value is about twice as large.

When a variance is essentially zero, `brsmm()` warns "Variance component at
the boundary" (log SD below $-6$, or a log-likelihood gain below $10^{-3}$
over the same fit without the term). If that gain is negative, the message
adds that SD $\approx 0$ has a higher log-likelihood: the fit stopped short
of the maximum. The Wald standard error of that log SD is meaningless; rely
on the likelihood-ratio test.

```{r lr-evolution}
# Base model without a random effect
fit_brs <- brs(
  y ~ x1,
  data = sim$data,
  repar = 2
)

# Reuse the mixed models already fitted:
# fit_mm    : random = ~ 1 | id
# fit_mm_rs : random = ~ 1 + x1 | id

tab_lr <- anova(fit_brs, fit_mm, fit_mm_rs, test = "Chisq")
tab_lr
```

Operational decision rule (analytical):

* If the second jump (RI to RI+RS) does not improve the fit (high p-value), prefer the random-intercept model for parsimony.
* If there is a robust gain, adopt the RI+RS model and validate parameter stability (especially `sd_b` and the $D$ matrix) via sensitivity and residual diagnostics.

### Residual diagnostics (quick checks)

```{r quick-diagnostics}
r <- residuals(fit_mm, type = "pearson")
kbl10(
  data.frame(
    mean = mean(r),
    sd = stats::sd(r),
    q025 = as.numeric(stats::quantile(r, 0.025)),
    q975 = as.numeric(stats::quantile(r, 0.975))
  ),
  digits = 4
)
```

## Parameter recovery experiment

A single-fit recovery table can be produced directly from the previous fit:

```{r recovery-single}
est <- c(
  beta0 = unname(coef(fit_mm, model = "mean")[1]),
  beta1 = unname(coef(fit_mm, model = "mean")[2]),
  sigma_b = unname(exp(coef(fit_mm, model = "random")))
)

true <- c(
  beta0 = sim$truth$beta[1],
  beta1 = sim$truth$beta[2],
  sigma_b = sim$truth$sigma_b
)

recovery_table <- data.frame(
  parameter = names(true),
  true = as.numeric(true),
  estimate = as.numeric(est[names(true)]),
  bias = as.numeric(est[names(true)] - true)
)
kbl10(recovery_table)
```

For a Monte Carlo recovery study, repeat simulation and fitting across
replicates:

```{r recovery-montecarlo, eval = FALSE}
mc_recovery <- function(R = 50L, seed = 7001L) {
  set.seed(seed)
  out <- vector("list", R)

  for (r in seq_len(R)) {
    sim_r <- sim_brsmm_data(seed = seed + r)
    fit_r <- brsmm(
      y ~ x1,
      random = ~ 1 | id,
      data = sim_r$data,
      repar = 2,
      int_method = "laplace",
      method = "BFGS",
      control = list(maxit = 1000)
    )

    out[[r]] <- c(
      beta0 = unname(coef(fit_r, model = "mean")[1]),
      beta1 = unname(coef(fit_r, model = "mean")[2]),
      sigma_b = unname(exp(coef(fit_r, model = "random")))
    )
  }

  est <- do.call(rbind, out)
  truth <- c(beta0 = 0.20, beta1 = 0.65, sigma_b = 0.55)

  data.frame(
    parameter = colnames(est),
    truth = as.numeric(truth[colnames(est)]),
    mean_est = colMeans(est),
    bias = colMeans(est) - truth[colnames(est)],
    rmse = sqrt(colMeans((sweep(est, 2, truth[colnames(est)], "-"))^2))
  )
}

kbl10(mc_recovery(R = 50))
```

## How this maps to automated package tests

The package test suite includes dedicated `brsmm` tests for:

1. fitting with Laplace integration;
2. one- and two-part formulas;
3. S3 methods (`coef`, `vcov`, `summary`, `predict`, `residuals`, `ranef`);
4. parameter recovery under known DGP settings.

Run locally:

```{r run-tests, eval = FALSE}
devtools::test(filter = "brsmm")
```

## References

Self, S. G., and Liang, K.-Y. (1987). Asymptotic properties of maximum
likelihood estimators and likelihood ratio tests under nonstandard
conditions. *Journal of the American Statistical Association*, 82(398),
605-610.

Stram, D. O., and Lee, J. W. (1994). Variance components testing in the
longitudinal mixed effects model. *Biometrics*, 50(4), 1171-1177.

Ferrari, S. L. P. and Cribari-Neto, F. (2004). Beta regression for modelling
rates and proportions. *Journal of Applied Statistics*, 31(7), 799-815.
DOI: 10.1080/0266476042000214501. Validated online via:
[https://doi.org/10.1080/0266476042000214501](https://doi.org/10.1080/0266476042000214501).

Pinheiro, J. C. and Bates, D. M. (2000). *Mixed-Effects Models in S and S-PLUS*.
Springer. DOI: 10.1007/b98882. Validated online via:
[https://doi.org/10.1007/b98882](https://doi.org/10.1007/b98882).

Rue, H., Martino, S., and Chopin, N. (2009). Approximate Bayesian inference
for latent Gaussian models by using integrated nested Laplace approximations.
*Journal of the Royal Statistical Society: Series B*, 71(2), 319-392.
DOI: 10.1111/j.1467-9868.2008.00700.x. Validated online via:
[https://doi.org/10.1111/j.1467-9868.2008.00700.x](https://doi.org/10.1111/j.1467-9868.2008.00700.x).
